---
title: "BNC5 Metabolomics Analysis — Pellet Positive Mode"
subtitle: "Response of Bacillus amyloliquefaciens BNC5 to Ralstonia solanacearum Whole Cells"
author: "Augustine Onuh, PhD | Chulalongkorn University"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE,
                      fig.width = 9, fig.height = 6)
suppressPackageStartupMessages({
  library(tidyverse)
  library(FactoMineR)
  library(factoextra)
  library(mixOmics)
  library(ggrepel)
  library(knitr)
  library(kableExtra)
})
load("BNC5_pellet_pos_analysis_session.RData")
```

## Executive Summary

This report presents untargeted LC-MS metabolomics analysis of
*Bacillus amyloliquefaciens* BNC5 pellet extracts comparing three
conditions: BNC5 alone, BNC5 co-cultured with *Ralstonia solanacearum*
whole cells (BNC5RSw), and RSw alone (dead cell background control).

**Key findings:**

- **`r sum(anova_results$Significant)`** features showed significant
  variation across groups (ANOVA, BH-adjusted p < 0.05)
- **`r nrow(elevated_in_treatment)`** features were elevated in BNC5RSw
  vs BNC5 (log2FC > 0.58, p_BH < 0.05)
- After removing **`r sum(elevated_in_treatment$Feature %in% rsw_background_features)`**
  RSw background features, **`r nrow(true_response)`** true BNC5
  response features were identified
- **`r nrow(true_response[true_response$raw_FC > 10,])`** features
  showed greater than 10-fold elevation in BNC5RSw vs BNC5
- **`r nrow(reduced_features)`** features were significantly reduced
  in BNC5RSw compared to BNC5 alone

## Study Design

```{r design-table}
design_summary <- data.frame(
  Condition = c("BNC5", "BNC5RSw", "RSw"),
  Description = c(
    "B. amyloliquefaciens BNC5 alone — negative control",
    "BNC5 co-cultured with R. solanacearum whole cells — treatment",
    "R. solanacearum whole cells alone — background control"
  ),
  Replicates = c(
    "6 (3 x Lot1 + 3 x Lot2)",
    "6 (3 x Lot1 + 3 x Lot2)",
    "6 (3 x Lot1 + 3 x Lot2)"
  ),
  Role = c(
    "Baseline BNC5 metabolome",
    "RSw-induced metabolic response",
    "Background subtraction control"
  )
)

kable(design_summary,
      caption = "Table 1: Experimental design and sample overview") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Instrument:** HPLC-QTOF ESI-MS (positive ionisation mode)
**Sample type:** Bacterial cell pellets
**Biological lots:** 2 independent preparations per condition
**Total samples analysed:** 18

## Data Processing

### XCMS Peak Detection

Raw mzXML files were processed using XCMS (v4.10.1):

- Peak detection: CentWave algorithm
- Retention time correction: LOESS smoothing, span = 0.6, 2 rounds
- Peak grouping: density method, bandwidth = 10
- Missing peak filling: fillPeaks()
- Adduct and isotope annotation: CAMERA (positive mode)
- **Total features detected: `r nrow(raw_data)`**

### Preprocessing Pipeline

```{r preprocessing-table}
preproc <- data.frame(
  Step = c(
    "Raw XCMS features",
    "Zeros converted to NA",
    "50% per-group presence filter",
    "Half-minimum imputation",
    "TIC normalisation",
    "Log2 transformation",
    "Autoscaling (z-score per feature)"
  ),
  Result = c(
    paste(nrow(raw_data), "features detected"),
    paste(sum(raw_data == 0, na.rm = TRUE), "zeros replaced with NA"),
    paste(nrow(data_filtered), "features retained — 0 removed"),
    "All missing values replaced — 0 NAs remaining",
    "All 18 samples normalised to equal total ion count",
    "Log-normal distribution corrected",
    "Mean = 0, SD = 1 per feature across samples"
  )
)

kable(preproc,
      caption = "Table 2: Step-by-step preprocessing summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## Quality Control

### Lot Effect Assessment

```{r lot-pca, fig.cap="Figure 1: PCA coloured by biological lot. Complete overlap between Lot1 and Lot2 confirms no significant technical batch effect between the two independent biological replicates."}
fviz_pca_ind(pca_result,
             geom.ind      = "point",
             col.ind       = lots,
             palette       = c("#ff7f0e", "#9467bd"),
             addEllipses   = TRUE,
             ellipse.level = 0.95,
             legend.title  = "Lot",
             title = "PCA — Lot Effect Assessment") +
  theme_bw(base_size = 12)
```

Lot1 and Lot2 samples overlapped completely in PCA space confirming
no significant technical effect between the two independent biological
lots. Lot was therefore not included as a covariate in statistical models.

### Biological Group Separation

```{r group-pca, fig.cap="Figure 2: PCA scores plot coloured by biological group. PC1 separates BNC5 from RSw-containing groups. PC2 separates BNC5RSw from RSw alone. All three groups show complete separation with no overlap between 95% confidence ellipses."}
fviz_pca_ind(pca_result,
             geom.ind      = "point",
             col.ind       = groups,
             palette       = c("#2ca02c", "#1f77b4", "#d62728"),
             addEllipses   = TRUE,
             ellipse.level = 0.95,
             legend.title  = "Group",
             title = paste0("PCA Scores Plot — PC1: ",
                            var_exp[1], "%, PC2: ",
                            var_exp[2], "%")) +
  theme_bw(base_size = 12)
```

PC1 (**`r var_exp[1]`%** of variance) clearly separated BNC5 from the
RSw-containing conditions. PC2 (**`r var_exp[2]`%**) distinguished
BNC5RSw from RSw alone. The positioning of BNC5RSw between BNC5 and
RSw along PC1 reflects its composite metabolome containing both
BNC5-derived and RSw-derived metabolites.

## Statistical Analysis

### ANOVA Results

**`r sum(anova_results$Significant)`** of `r nrow(anova_results)`
features showed significant variation across the three groups after
Benjamini-Hochberg correction (p < 0.05).

```{r anova-table}
anova_top15 <- head(anova_results, 15) %>%
  mutate(
    ANOVA_p    = formatC(ANOVA_p,    format = "e", digits = 2),
    ANOVA_p_BH = formatC(ANOVA_p_BH, format = "e", digits = 2)
  )

kable(anova_top15,
      caption = "Table 3: Top 15 features by ANOVA significance (BH-corrected)",
      col.names = c("Feature", "Raw p-value",
                    "BH-adjusted p", "Significant"),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### RSw Background Filter

To identify metabolites specifically produced by BNC5 in response to
RSw, a three-step filter was applied:

| Filter Step | Count |
|---|---|
| Features elevated in BNC5RSw vs BNC5 (log2FC > 0.58, p_BH < 0.05) | `r nrow(elevated_in_treatment)` |
| Removed — also elevated in RSw background | `r sum(elevated_in_treatment$Feature %in% rsw_background_features)` |
| **True BNC5 response features** | **`r nrow(true_response)`** |
| Features reduced in BNC5RSw vs BNC5 | `r nrow(reduced_features)` |

### Volcano Plot

```{r volcano, fig.width=10, fig.height=7, fig.cap="Figure 3: Volcano plot of BNC5RSw vs BNC5. Red = true BNC5 response features. Orange = RSw background features removed by filter. Blue = features reduced in BNC5RSw. Grey = not significant."}
ggplot(volcano_data,
       aes(x = log2FC, y = neg_log10p,
           colour = Direction, label = Label)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_text_repel(size = 2.8, max.overlaps = 20,
                  colour = "black", box.padding = 0.3) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(-0.58, 0.58),
             linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(
    "True BNC5 Response" = "#d62728",
    "RSw Background"     = "#ff7f0e",
    "Reduced in BNC5RSw" = "#1f77b4",
    "Not Significant"    = "grey70"
  )) +
  labs(
    x      = "Log2 Fold Change (BNC5RSw vs BNC5)",
    y      = "-log10(BH-adjusted p-value)",
    colour = "Category"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Multivariate Analysis — PLS-DA

```{r plsda-plot, fig.cap="Figure 4: PLS-DA scores plot using 1,533 true BNC5 response features. Complete separation of all three groups confirms biological specificity of the identified response metabolome."}
plotIndiv(plsda_model,
          comp      = c(1, 2),
          group     = as.factor(groups),
          ind.names = FALSE,
          ellipse   = TRUE,
          legend    = TRUE,
          title     = "PLS-DA: True BNC5 Response Features",
          col       = c("#2ca02c", "#1f77b4", "#d62728"))
```

## Top BNC5 Response Features

```{r top-features}
top20 <- head(master_table, 20) %>%
  mutate(p_BH = formatC(p_BH, format = "e", digits = 2))

kable(top20,
      caption = "Table 4: Top 20 true BNC5 response features ranked by VIP score",
      col.names = c("Feature", "m/z", "RT (min)",
                    "Log2FC", "Fold Change", "p_BH", "VIP Score"),
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 11) %>%
  column_spec(5, bold = TRUE, color = "#d62728")
```

## Conclusions

Untargeted positive-mode LC-MS metabolomics of *B. amyloliquefaciens*
BNC5 pellet extracts identified **`r nrow(true_response)`** features
specifically elevated in BNC5 upon exposure to *R. solanacearum* whole
cells, after rigorous removal of RSw background metabolites
(n = `r sum(elevated_in_treatment$Feature %in% rsw_background_features)`).

A further **`r nrow(reduced_features)`** features were significantly
reduced in BNC5RSw compared to BNC5 alone, indicating active metabolic
reprogramming in response to pathogen challenge — consistent with
resource reallocation toward defensive compound biosynthesis.

The complete absence of lot effect confirms biological reproducibility
across two independent experimental preparations. PCA and PLS-DA both
confirmed complete metabolic separation between all three groups,
validating the experimental design and computational workflow.

**Recommended next steps:**

1. Putative annotation of the `r nrow(true_response)` true BNC5
   response features using accurate mass database searching
   (HMDB, METLIN, LipidMaps) and MS/MS fragmentation confirmation
2. Supernatant fraction analysis using the identical pipeline to
   characterise secreted metabolites
3. Negative ionisation mode analysis for complementary lipid and
   organic acid coverage

---
*Analysis performed by Augustine Onuh, PhD*  
*Department of Chemistry, Chulalongkorn University, Bangkok, Thailand*  
*Generated: `r Sys.Date()` | R `r getRversion()` | xcms v4.10.1*